---
title: "193DS_spring_2025_final"
---

https://github.com/Yonatan-Grossman/ENVS-193DS_spring-2025_final

# Problem 1

### a.

My coworker used a Pearson's r test in part 1, as it tests for corrilation. They used a one-way analysis of variance (ANOVA) in part 2 since they were checking to see if there was a difference in the means of multiple groups.

### b.

Two things they should have included are the degrees of freedom and the F-statistic. The F-statistic is the main output of an ANOVA, and it shows the ratio of variance between group means to the variance within group means, so showing the F-statistic allows the reader to better understand the difference between group means. In this case, since the test rejected the null hypothesis, the F-statistic would probably be high and indicate that the difference in the means of each of the groups (urban land, etc) was larger than the difference within each of those groups, with a higher F-statistic meaning the average nitrogen load varies much less within groups compared to how those groups compare to each other. Knowing the F-statistic the better we can understand how different the means of the groups are. The degrees of freedom allow the reader to better understand how much information was used in the ANOVA and know the rough shape of the F-distribution. In this case, a higher df means more of each source was sampled, so they would have more data points to work with.

### c.

We found a statistically significant difference between average nitrogen load (kg year\^-1) between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) (one-way ANOVA, F(df, df) = F-statistic, p = 0.02, α = significance level).

We also found a \[+/-, strength\] correlation between distance from headwater (km) and annual total nitrogen load (kg year\^-1) (Pearson's r(df) = correlation coefficient, p=0.02 α = significance level)

# Problem 2.

### a.

```{r}

library(tidyverse)
library(janitor)

```

```{r}

sst <- read.csv("data/SST_update2023.csv")

sst_clean <- sst %>%
  clean_names() %>%
  mutate(
    year = year(date),
    month = month(date, label = TRUE, abbr = TRUE)) %>% 
  filter(year > 2017) %>% 
  mutate(year = factor(year)) %>% 
  group_by(year, month) %>% 
  summarize(
    mean_monthly_sst = mean(temp, na.rm = TRUE),
    .groups = "drop"
  )


ggplot(sst_clean,
       aes(x = month, 
           y = mean_monthly_sst,
           color = year,
           group = year))+
  geom_line() + 
  geom_point() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

